---
title: "Homework #3: Penalized Regression" 
author: "Aubrey Winger"
date: "Due: Wed Sept 21 | 11:45am"
output: R6030::homework
---

**DS 6030 | Fall 2022 | University of Virginia**

------------------------------------------------------------------------

```{r config, echo=FALSE}
source(system.file("config/hw_config.R", package="R6030")) # knitr settings
# options(dplyr.summarise.inform = FALSE)  # ignore dplyr message about grouping
```


# Required R packages and Directories

::: {.solution}
```{r packages, message=FALSE, warning=FALSE}
data.dir = 'https://mdporter.github.io/DS6030/data/' # data directory
library(mlbench)
library(glmnet)
library(R6030)     # functions for DS-6030
library(tidyverse) # functions for data manipulation   
```
:::


# Problem 1: Optimal Tuning Parameters

In cross-validation, we discussed choosing the tuning parameter values that minimized the cross-validation error. Another approach, called the "one-standard error" rule [ISL pg 214, ESL pg 61], uses the values corresponding to the least complex model whose cv error is within one standard error of the best model. The goal of this assignment is to compare these two rules.

Use simulated data from `mlbench.friedman1(n, sd=2)` in the `mlbench` R package to fit *lasso models*. The tuning parameter $\lambda$ (corresponding to the penalty on the coefficient magnitude) is the one we will focus one. Generate training data, use k-fold cross-validation to get $\lambda_{\rm min}$ and $\lambda_{\rm 1SE}$, generate test data, make predictions for the test data, and compare performance of the two rules under a squared error loss using a hypothesis test.


Choose reasonable values for:

- Number of cv folds ($K$)
    - Note: you are free to use repeated CV, repeated hold-outs, or bootstrapping instead of plain cross-validation; just be sure to describe what do did so it will be easier to follow.
- Number of training and test observations
- Number of simulations
- If everyone uses different values, we will be able to see how the results change over the different settings.
- Don't forget to make your results reproducible (e.g., set seed)

This pseudo code will get you started:
```yaml
library(mlbench)
library(glmnet)

#-- Settings
n.train =        # number of training obs
n.test =         # number of test obs
K =              # number of CV folds
alpha =          # glmnet tuning alpha (1 = lasso, 0 = ridge)
M =              # number of simulations

#-- Data Generating Function
getData <- function(n) mlbench.friedman1(n, sd=2) # data generating function

#-- Simulations
# Set Seed Here

for(m in 1:M) {

# 1. Generate Training Data
# 2. Build Training Models using cross-validation, e.g., cv.glmnet()
# 3. get lambda that minimizes cv error and 1 SE rule
# 4. Generate Test Data
# 5. Predict y values for test data (for each model: min, 1SE)
# 6. Evaluate predictions

}

#-- Compare
# compare performance of the approaches / Statistical Test
```

## a. Code for the simulation and performance results

::: {.solution}
Lasso regression
```{r}

#-- Settings
n.train =  800      # number of training obs
n.test =   300      # number of test obs
K =    10          # number of CV folds
alpha =   1     # glmnet tuning alpha (1 = lasso, 0 = ridge)
M =    30          # number of simulations

#-- Data Generating Function
getData <- function(n) mlbench.friedman1(n, sd=2) # data generating function

#-- Simulations
# Set Seed Here
MSE_min <- rep(0,M)
MSE_1se <- rep(0,M)
set.seed(201)
for(m in 1:M) {
  train_data <- getData(n.train)
  ridge_cv = cv.glmnet(train_data$x,train_data$y,alpha=alpha,
                    nfolds=K)
  test_data <- getData(n.test)
  yhat_1se = predict(ridge_cv, test_data$x, s = "lambda.1se")
  yhat_min = predict(ridge_cv, test_data$x, s="lambda.min")
  MSE_1se[m] <- mean((test_data$y-yhat_1se)^2) #is this the same as squared error loss?
  MSE_min[m] <- mean((test_data$y-yhat_min)^2)
# 1. Generate Training Data
# 2. Build Training Models using cross-validation, e.g., cv.glmnet()
# 3. get lambda that minimizes cv error and 1 SE rule
# 4. Generate Test Data
# 5. Predict y values for test data (for each model: min, 1SE)
# 6. Evaluate predictions

}
lasso = data.frame(MSE_min,MSE_1se)
```
:::


## b. Description and results of a hypothesis test comparing $\lambda_{\rm min}$ and $\lambda_{\rm 1SE}$.

::: {.solution}
I conducted a two sided paired t-test since the MSE values for the two different lambda calculations come from the same distribution. I wanted to test if the performance of the two rules under squared error was significantly different from each other. My null hypothesis is that the true difference between the means of the squared error losses is equal to zero, meaning that the two lambda calculations do not result in significantly different losses. My alternative hypothesis is that the true difference between the means of the squared error losses is not equal to zero, meaning that the two lambda calculations do have significantly different losses. The resulting p value is significant at an alpha level of 0.05, meaning the null hypothesis is rejected. It can be concluded that there is a significant difference between the losses for lambda that minimizes SV and lambda that minimizes the 1 SE rule.
```{r}
#what type of hypothesis test
t.test(lasso$MSE_min,lasso$MSE_1se,paired=TRUE,alternative="two.sided")
```

:::

# Problem 2 Prediction Contest: Real Estate Pricing

This problem uses the [realestate-train](`r file.path(data.dir, 'realestate-train.csv')`) and [realestate-test](`r file.path(data.dir, 'realestate-test.csv')`) (click on links for data). 

The goal of this contest is to predict sale price (in thousands) (`price` column) using an *elastic net* model. Evaluation of the test data will be based on the root mean squared error ${\rm RMSE}= \sqrt{\frac{1}{m}\sum_i (y_i - \hat{y}_i)^2}$ for the $m$ test set observations. 


## a. Load the data and create necessary data structures for running *elastic net*.
- You are free to use any data transformation or feature engineering
- Note: there are some categorical predictors so at the least you will have to convert those to something numeric (e.g., one-hot or dummy coding). 

::: {.solution}
```{r}
test_data <- read_csv("realestate-test.csv",show_col_types = FALSE)
train_data <- read_csv("realestate-train.csv",show_col_types = FALSE)
```

:::


## b. Use an *elastic net* model to predict the `price` of the test data.  
- You are free to use any data transformation or feature engineering
- You are free to use any tuning parameters
- Report the $\alpha$ and $\lambda$ parameters you used to make your final predictions.
- Describe how you choose those tuning parameters

::: {.solution}
I chose the alpha value by iterating through 10 alpha values from 0 to 1 (incrementing by intervals of 0.1), and selecting the one that resulted in the best RMSE. In this case, it was alpha=0.8. I chose the lambda value by finding the minimum lambda value and using that in the predict function. As shown in the problem above, the minimum lambda results in a smaller loss than the lambda that minimizes the 1 SD rule, therefore this parameter was chosen. The lambda value I used was 0.5972. 
```{r}
set.seed(190)
sample <- sample(c(TRUE, FALSE), nrow(train_data), replace=TRUE, prob=c(0.8,0.2))
train_data$sample <- sample
X = glmnet::makeX(
train = train_data %>% filter(sample) %>% select(-price, -sample),
test = train_data %>% filter(!sample) %>% select(-price, -sample),
na.impute = TRUE
)
X.train = X$x
Y.train = train_data %>% filter(sample) %>% pull(price)
X.test = X$xtest
Y.test = train_data %>% filter(!sample) %>% pull(price)
```

```{r}
X_real_test = glmnet::makeX(
  train=test_data,
  na.impute = TRUE
)
```


```{r}
alpha = seq(from=0,to=1,by=0.1)
for (i in alpha){
model_cv = cv.glmnet(X.train,Y.train,alpha=i,
                    nfolds=10, standardize=TRUE)
  yhat_est = predict(model_cv, X.test, s=model_cv$lambda.min)
  RMSE <- sqrt(mean((Y.test-yhat_est)^2))
  print(RMSE)
}
```

```{r}
  #predict with the unlabeled test data given
lasso_cv = cv.glmnet(X.train,Y.train,alpha=0.8,
                    nfolds=10, standardize=TRUE)
yhat = predict(lasso_cv,X_real_test,s="lambda.min")
```

```{r}
lasso_cv$lambda.min
```

:::

## c. Submit a .csv file (ensure comma separated format) named `lastname_firstname.csv` that includes the column named *yhat* that is your estimates. We will use automated evaluation, so the format must be exact.  
- You will receive credit for a proper submission; the top five scores will receive 2 bonus points.     
    
::: {.solution}
```{r}
colnames(yhat)<-c("yhat") #need to remove the index?
final <- write.csv(yhat,"winger_aubrey.csv",row.names = FALSE)
```

:::

## d. Report the anticipated performance of your method in terms of RMSE. We will see how close your performance assessment matches the actual value. 

::: {.solution}
The RMSE achieved on the training data was 32.43, therefore this is the anticipated performance. 
:::   
